---
output:
  html_document:
    css: www/styles.css
    anchor_sections: FALSE
params:
  file: "MIDN_tree_sapling_data.csv"
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = FALSE, vline = "|")
options(knitr.kable.NA = '', scipen = 100)
```

```{r, results = 'hide', warning=FALSE, message=FALSE}
library(tidyverse)
#library(rstatix) # for outlier detection
library(knitr)
library(kableExtra)
#dat <- read.csv("./data/MIDN_tree_sapling_data.csv")
dat <- read.csv(paste0("./data/", params$file))
```

## MIDN QC {.tabset}

### Tree Heights {.tabset}
Checking height codes that vary more than 10% among years within the same plot. MIDN methods changed over time (see below), so a bit tricky to QC. 

History of tree height in MIDN:
* 2007: tree height not sampled- used 2011 values
* 2008: Only measured height of one codominant tree per plot, and didn't record tree tag.
* 2009-2010: Started measuring 3 codominant trees per plot, but didn't record tree tags.
* 2011: Began recording tree tag of measured trees. 
* 2013 and later: Started measuring tree height on 3 intermediate trees in addition to 3 codominant trees. 

There aren't nearly as many rogue values as NETN. Wondering if it's because MIDN used hypsometer from the beginning, and possibly that canopy heights are more uniform across plots/parks (ie crews get used to recording a range of numbers). Heights from the first cycle do tend to be the cycle with outliers, like with NETN.

```{r results = 'hide'}
hts <- 
  rbind(dat |> filter(year > 2007 & year < 2011) |> filter(treeclcd == 3) |> select(park, plt_cn, year, cycle, treeclcd, ht) |> unique(), # early data
        dat |> filter(year >= 2011) |> filter(htcd == 1) |> select(park, plt_cn, year, cycle, treeclcd, ht) |> unique() # latest protocol data
        ) |> filter(!is.na(ht)) |> filter(treeclcd %in% c(3, 4))

# hts_out <- hts |> group_by(park, plt_cn, treeclcd) |> identify_outliers(ht) |> 
#   mutate(outlier_cycle = cycle) |> filter(is.extreme == TRUE) |> 
#   select(park, plt_cn, treeclcd, cycle, ht, outlier_cycle) 

table(dat$htcd, dat$treeclcd)

hts_out <- hts |> group_by(park, plt_cn, treeclcd) |> 
  summarize(ht_max = quantile(ht, probs = 0.99, na.rm = T), 
            ht_min = quantile(ht, probs = 0.01, na.rm = T), 
            .groups = 'drop')

hts_sum <- hts |> group_by(park, plt_cn, cycle, treeclcd) |> 
  filter(treeclcd %in% c(3,4)) |> # remove a couple rogue tree heights for other crown classes
  summarize(hts_avg = round(mean(ht), 2), .groups = 'drop')

hts_join <- left_join(hts_sum, hts_out, by = c("park", "plt_cn", "treeclcd"))
head(hts_join)

table(hts_join$cycle)

hts_wide <- hts_join |> unique() |> 
  arrange(plt_cn, cycle, treeclcd) |> 
  pivot_wider(names_from = cycle, 
              values_from = hts_avg, 
              names_prefix = "c") |> 
  select(park, plt_cn, treeclcd, ht_max, ht_min, c1, c2, c3, c4, c5) |> 
  mutate(across(c1:c5, 
                ~ifelse(.x < ht_min, "low", ifelse(.x > ht_max, "high", NA_character_)),
                .names = "{.col}_out")) |> 
  filter(if_any(c1_out:c5_out, ~!is.na(.))) |> 
  mutate(crown_class = case_when(treeclcd == 3 ~ 'codom',
                                 treeclcd == 4 ~ 'inter',
                                 TRUE ~ NA_character_)) |> 
  arrange(crown_class, plt_cn) 

head(hts_wide)
# no outliears in c2-4
ht_kbl <- kable(hts_wide |> select(plt_cn, crown_class, c1:c5), format = 'html', align = 'c',
                col.names = c("Plot", "Crown Class", paste0("C", 1:5, " height")),
                caption = "Tree heights outside of 99% quantile within a plot and crown class. Values are tree heights measured in meters. Red are high and blue are low outliers") |> 
  kable_styling(fixed_thead = T, bootstrap_options = c('condensed'), full_width = F, position = 'center') |> 
  column_spec(1:7, border_left = "1px solid grey", border_right = "1px solid grey") |> 
  column_spec(3, background = ifelse(hts_wide$c1_out %in% "high", "#eaafaf",
                                     ifelse(hts_wide$c1_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(4, background = ifelse(hts_wide$c2_out %in% "high", "#eaafaf",
                                     ifelse(hts_wide$c2_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(5, background = ifelse(hts_wide$c3_out %in% "high", "#eaafaf",
                                     ifelse(hts_wide$c3_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(6, background = ifelse(hts_wide$c4_out %in% "high", "#eaafaf",
                                     ifelse(hts_wide$c4_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(7, background = ifelse(hts_wide$c5_out %in% "high", "#eaafaf",
                                     ifelse(hts_wide$c5_out %in% "low", "#AFC0EA", "white")))
```

```{r include = F}
head(hts_join)
hts_out2 <- hts_join |> mutate(outlier = ifelse(hts_avg > ht_max, "high", ifelse(hts_avg < ht_min, "low", NA_character_))) |> 
  filter(treeclcd == 3)
head(hts_wide)
codom_outs <- unique(hts_wide$plt_cn[hts_wide$crown_class == "codom"])

head(hts_out2)
midn <- c("APCO", "BOWA", "FRSP", "GETT", "HOFU", "PETE", "RICH", "VAFO")
midn <- ggplot(hts_out2 |> filter(plt_cn %in% codom_outs) |> filter(park %in% midn), 
               aes(x = cycle, y = hts_avg)) +
        geom_line() + facet_wrap(~plt_cn) + theme_bw() +
        labs(x = "Cycle", y = "Tree Ht (m)") +
        theme(strip.text = element_text(size = 7))

ncbn <- ggplot(hts_out2 |> filter(plt_cn %in% codom_outs) |> filter(!park %in% midn), 
               aes(x = cycle, y = hts_avg)) +
       geom_line() + facet_wrap(~plt_cn) + theme_bw() +
        labs(x = "Cycle", y = "Tree Ht (m)") +
        theme(strip.text = element_text(size = 7))

```


```{r results = 'hide'}
#Trying to better identify the big swings by comparing value to +/- 25% of median
hts_med <- hts |> group_by(park, plt_cn, treeclcd) |> 
  summarize(ht_med = median(ht, na.rm = T),
            ht_p25 = ht_med*1.25,
            ht_m25 = ht_med*0.25, 
            .groups = 'drop')

hts_join2 <- left_join(hts_sum, hts_med, by = c("park", "plt_cn", "treeclcd")) 
head(hts_join2)

hts_wide2 <- hts_join2 |> unique() |> 
  pivot_wider(names_from = cycle, 
              values_from = hts_avg, 
              names_prefix = "c") |> 
  arrange(plt_cn, treeclcd) |> 
  mutate(across(c1:c5, 
                ~ifelse(.x < ht_m25, "low", ifelse(.x > ht_p25, "high", NA_character_)),
                .names = "{.col}_out")) |> 
  filter(if_any(c1_out:c5_out, ~!is.na(.))) |> 
  mutate(crown_class = case_when(treeclcd == 3 ~ 'codom',
                                 treeclcd == 4 ~ 'inter',
                                 TRUE ~ NA_character_)) |> 
  arrange(crown_class, plt_cn)

head(hts_wide2)

codom_outs2 <- unique(hts_wide2$plt_cn[hts_wide2$crown_class == "codom"])

head(hts_out2)

outs50 <- ggplot(hts_out2 |> filter(plt_cn %in% codom_outs2), 
                 aes(x = cycle, y = hts_avg)) +
          geom_line() + facet_wrap(~plt_cn) + theme_bw() +
          labs(x = "Cycle", y = "Tree Ht (m)") +
          theme(strip.text = element_text(size = 7))

```

#### Outliers +/- 25%
MIDN/NCBN plots with at least 1 identified outlier (+/- 25% of median) for codominant trees. 
```{r fig.height = 10, out.width = '100%'}
outs50
```

#### Outlier Table

```{r}
ht_kbl
```


#### MIDN less extreme outs
MIDN plots with at least 1 identified outlier for codominant trees as being outside 99% quantile of heights measured within a given plot. 
```{r fig.height = 10, out.width = '100%'}
midn
```

#### NCBN less extreme outs
NCBN plots with at least 1 identified outlier for codominant trees as being outside 99% quantile of heights measured within a given plot.
```{r fig.height = 10, out.width = '100%'}
ncbn
```

### Tree measurements
#### Swings in DBH
```{r results = 'hide'}
dbh1 <- dat |> select(park, plt_cn, tre_cn, cycle, treeclcd, dbhcm) |> 
  pivot_wider(names_from = cycle, values_from = dbhcm, names_prefix = "c")

dbhstat <- dat |> select(park, plt_cn, tre_cn, cycle, treeclcd, statuscd) |> 
  pivot_wider(names_from = cycle, values_from = statuscd, names_prefix = "status_c")

dbh <- full_join(dbh1, dbhstat, by = c("park", "plt_cn", "tre_cn", "treeclcd"))

dbh$c12_dif = dbh$c2 - dbh$c1
dbh$c23_dif = dbh$c3 - dbh$c2
dbh$c34_dif = dbh$c4 - dbh$c3
dbh$c45_dif = dbh$c5 - dbh$c4

dbh$c12_out = ifelse(dbh$c12_dif > 4, "high", ifelse(dbh$c12_dif < -0.5, "low", NA_character_))
dbh$c23_out = ifelse(dbh$c23_dif > 4, "high", ifelse(dbh$c23_dif < -0.5, "low", NA_character_))
dbh$c34_out = ifelse(dbh$c34_dif > 4, "high", ifelse(dbh$c34_dif < -0.5, "low", NA_character_))
dbh$c45_out = ifelse(dbh$c45_dif > 4, "high", ifelse(dbh$c45_dif < -0.5, "low", NA_character_))

dbh_out <- dbh |> filter(if_any(c(c12_out, c23_out, c34_out, c45_out), ~!is.na(.))) |> data.frame() |> 
  filter(!if_any(c(status_c1, status_c2, status_c3, status_c4, status_c5), ~.x=='dead'))
head(dbh_out)

dbh_kbl <- kable(dbh_out |> select(plt_cn, treeclcd, c1:c5), format = 'html', align = 'c',
                col.names = c("Plot", "Crown Class", paste0("C", 1:5, " height")),
                caption = "DBH measurements >4cm DBH or <-0.5cm DBH of previous measurement for live trees only. Red are high and blue are low outliers") |> 
  kable_styling(fixed_thead = T, bootstrap_options = c('condensed'), full_width = F, position = 'center') |> 
  column_spec(1:7, border_left = "1px solid grey", border_right = "1px solid grey") |> 
  column_spec(4, background = ifelse(dbh_out$c12_out %in% "high", "#eaafaf",
                                     ifelse(dbh_out$c12_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(5, background = ifelse(dbh_out$c23_out %in% "high", "#eaafaf",
                                     ifelse(dbh_out$c23_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(6, background = ifelse(dbh_out$c34_out %in% "high", "#eaafaf",
                                     ifelse(dbh_out$c34_out %in% "low", "#AFC0EA", "white"))) |> 
  column_spec(7, background = ifelse(dbh_out$c45_out %in% "high", "#eaafaf",
                                     ifelse(dbh_out$c45_out %in% "low", "#AFC0EA", "white"))) 
```


```{r}
dbh_kbl
```


#### Zombie trees

```{r}
head(dbh_out)
dbh_zomb <- dbh |> select(park, plt_cn, tre_cn, status_c1, status_c2, status_c3, status_c4, status_c5) |> 
  mutate(zombie = ifelse(status_c1 == "dead" & status_c2 == "live" | 
                         status_c2 == "dead" & status_c3 == "live" | 
                         status_c3 == "dead" & status_c4 == "live" |
                         status_c4 == "dead" & status_c5 == "live", 
                         "zombie", NA_character_)) |> 
  filter(zombie == "zombie")

```

#### Spp Changes
